---
title: "Example University DB Implementation"
author: "Raphael Russo"
output: html_notebook
---

## Overview
This is a simple example implementation of a relational database for a university based off of the schema outlined in `University_ERD.pdf` which is in BCNF. It includes students, majors, courses, faculty, etc. Data will be loaded a csv file of students and the courses they have taken, and stored in a SQLite database using R and SQL, and at the end, various queries will be made. Student data is generated using `generate_data.py` and can be found in `generated_u_data.csv`.

## Connect To Database

First, connect to the database, or create a new one if it does not already exist. 
```{r}
library(RSQLite)
library(DBI)
library(sqldf)
library(reticulate)
data.file <- "generated_u_data.csv"
dbfile <- "UniDB.db"
if (file.exists(dbfile)) {
  file.remove(dbfile) # just get rid of the db instead of dropping all the tables to start fresh
}
if (file.exists(data.file)) {
  file.remove(data.file)
}
py_run_file("generate_data.py")
dbcon <- dbConnect(RSQLite::SQLite(), dbfile)
dbExecute(dbcon, "PRAGMA foreign_keys = ON")
```

## Create Tables

Students Table:
```{sql createStudents, connection=dbcon}
CREATE TABLE IF NOT EXISTS Students (
  student_id INTEGER,
  first_name TEXT NOT NULL,
  last_name TEXT,
  birth_date DATE NOT NULL,
  email TEXT,
  expected_graduation DATE,
  PRIMARY KEY (student_id)
)
```

Faculty Table:
```{sql createFaculty, connection=dbcon}
CREATE TABLE IF NOT EXISTS Faculty (
  faculty_id INTEGER,
  first_name TEXT NOT NULL,
  last_name TEXT,
  birth_date DATE NOT NULL,
  email TEXT,
  PRIMARY KEY (faculty_id)
)
```

Departments Table:
TO DO - ADD DEPARTMENT HEAD IN DATA GEN
```{sql createDepartments, connection=dbcon}
CREATE TABLE IF NOT EXISTS Departments (
  department_id INTEGER,
  department_name TEXT NOT NULL,
  department_head INTEGER,
  department_code TEXT NOT NULL,
  PRIMARY KEY (department_id),
  FOREIGN KEY (department_head) REFERENCES Faculty (faculty_id) 
)
```

Department Membership Table:
```{sql createDepartmentMembers, connection=dbcon}
CREATE TABLE IF NOT EXISTS DepartmentMembership (
  dm_id INTEGER,
  faculty_id INTEGER NOT NULL,
  department_id INTEGER NOT NULL,
  PRIMARY KEY (dm_id),
  FOREIGN KEY (faculty_id) REFERENCES Faculty (faculty_id),
  FOREIGN KEY (department_id) REFERENCES Departments (department_id)
)
```

Courses Table:
```{sql createCourses, connection=dbcon}
CREATE TABLE IF NOT EXISTS Courses (
  course_id INTEGER,
  course_name TEXT NOT NULL,
  course_code INTEGER NOT NULL,
  department_id INTEGER NOT NULL,
  faculty_id INTEGER NOT NULL,
  semester_offered TEXT NOT NULL,
  PRIMARY KEY (course_id),
  FOREIGN KEY (department_id) REFERENCES Departments,
  FOREIGN KEY (faculty_id) REFERENCES Faculty (faculty_id)
)
```

Composite Key for Courses:
```{sql createCompositeCourseKeyView, connection=dbcon}
CREATE VIEW CoursesWithComposite AS
SELECT d.department_code || c.course_code || c.semester_offered AS composite_key
FROM Courses c, Departments d where c.department_id = d.department_id
```

Majors Table:
```{sql createMajors, connection=dbcon}
CREATE TABLE IF NOT EXISTS Majors (
  major_id INTEGER,
  major_name TEXT NOT NULL,
  department_id INTEGER NOT NULL,
  PRIMARY KEY (major_id),
  FOREIGN KEY (department_id) REFERENCES Departments (department_id)
)
```

Student Majors Table:
```{sql creatStudentMajors, connection=dbcon}
CREATE TABLE IF NOT EXISTS StudentMajors (
  student_id INTEGER NOT NULL,
  major_id INTEGER NOT NULL,
  CONSTRAINT student_major_id PRIMARY KEY (student_id, major_id),
  FOREIGN KEY (student_id) REFERENCES Students (student_id),
  FOREIGN KEY (major_id) REFERENCES Majors (major_id)
  
)
```


Required Courses Table:
```{sql createRequiredCourses, connection=dbcon}
CREATE TABLE IF NOT EXISTS RequiredCourses (
  rc_id INTEGER,
  course_id INTEGER NOT NULL,
  major_id INTEGER NOT NULL,
  PRIMARY KEY (rc_id),
  FOREIGN KEY (course_id) REFERENCES Courses (course_id),
  FOREIGN KEY (major_id) REFERENCES Majors (major_id)
)
```

Elective Courses Table:
```{sql createElectiveCourses, connection=dbcon}
CREATE TABLE IF NOT EXISTS ElectiveCourses (
  ec_id INTEGER,
  course_id INTEGER NOT NULL,
  major_id INTEGER NOT NULL,
  PRIMARY KEY (ec_id),
  FOREIGN KEY (course_id) REFERENCES Courses (course_id),
  FOREIGN KEY (major_id) REFERENCES Majors (major_id)
)
```

Classrooms Table:
```{sql createClassrooms, connection=dbcon}
CREATE TABLE IF NOT EXISTS Classrooms (
  classroom_id INTEGER,
  room_number INTEGER,
  building_name TEXT,
  capacity INTEGER,
  PRIMARY KEY (classroom_id)
)
```

Course Schedules Table:
```{sql createCourseSchedules, connection=dbcon}
CREATE TABLE IF NOT EXISTS CourseSchedules (
  schedule_id INTEGER,
  course_id INTEGER NOT NULL,
  classroom_id INTEGER NOT NULL,
  start_time TIME,
  end_time TIME,
  PRIMARY KEY (schedule_id)
)
```

Days of Week Table:
```{sql createDaysOfWeek, connection=dbcon}
CREATE TABLE IF NOT EXISTS DaysOfWeek (
    day_id INTEGER PRIMARY KEY,
    day_name TEXT(1) NOT NULL
);

```

Course Scheduled Days Table:
```{sql createScheduledDays, connection=dbcon}
CREATE TABLE IF NOT EXISTS CourseScheduleDays (
    schedule_id INTEGER,
    day_id INTEGER,
    CONSTRAINT cs_id PRIMARY KEY (schedule_id, day_id),
    FOREIGN KEY (schedule_id) REFERENCES CourseSchedules(schedule_id),
    FOREIGN KEY (day_id) REFERENCES DaysOfWeek(day_id)
);
```

Enrollment Table:
```{sql createEnrollmentTable, connection=dbcon}
CREATE TABLE IF NOT EXISTS Enrollment (
  student_id INTEGER,
  course_id INTEGER,
  grade NUMERIC,
  CONSTRAINT enrollment_id PRIMARY KEY (student_id, course_id)
)
```

## Check Table Creation

## Loading CSV Data

Read the student data csv:
```{r loadStudentData}
stu.data.raw <- read.csv("generated_u_data.csv")
```

Populate Days of Week table:
```{r insertDaysOfWeek}
days = c("U", "M", "T", "W", "R", "F", "S")
ndays = length(days)
sql <- "insert into DaysOfWeek (day_id, day_name) values"
for (i in 1:ndays) {
  day = days[i]
  val = paste0("(", i, ",'", day, "')")
  if (i != ndays) {
    val = paste0(val, ",")
  }
  sql <- paste0(sql, val)
}

print(sql)
rs <- dbExecute(dbcon, sql)
```
```{sql connection=dbcon}
SELECT * FROM DaysOfWeek
```
Populate Student Table
```{r}
get.students <- "select distinct StudentID,StudentFirstName,StudentLastName,StudentEmail,StudentDOB,StudentGradDate from `stu.data.raw`"
stu.info <- sqldf(get.students)
nrows <- nrow(stu.info)
sql <- "insert into Students (student_id, first_name, last_name, birth_date, email, expected_graduation) values"

for (i in 1:nrows) {
  id <- stu.info$StudentID[i]
  f.name <- stu.info$StudentFirstName[i]
  l.name <- stu.info$StudentLastName[i]
  email <- stu.info$StudentEmail[i]
  dob <- stu.info$StudentDOB[i]
  grad <- stu.info$StudentGradDate[i]
  val = paste0("(", id, ",'", f.name, "','", l.name, "','", dob, "','", email, "','", grad, "')")
  if (i != nrows) {
    val = paste0(val, ",")
  }
  sql <- paste0(sql, val)
}

rs <- dbExecute(dbcon, sql)
```
```{sql connection=dbcon}
SELECT * FROM STUDENTS
```
Populate Faculty Table
```{r}
get.fac <- "select distinct FacultyID,FacultyFirstName,FacultyLastName,FacultyDOB, FacultyEmail from `stu.data.raw`"
fac.info <- sqldf(get.fac)
nrows <- nrow(fac.info)
sql <- "insert into Faculty (faculty_id, first_name, last_name, birth_date, email) values"


for (i in 1:nrows) {
  id <- fac.info$FacultyID[i]
  f.name <- fac.info$FacultyFirstName[i]
  l.name <- fac.info$FacultyLastName[i]
  email <- fac.info$FacultyEmail[i]
  dob <- fac.info$FacultyDOB[i]
  val = paste0("(", id, ",'", f.name, "','", l.name, "','", dob, "','", email, "')")
  if (i != nrows) {
    val = paste0(val, ",")
  }
  sql <- paste0(sql, val)
}
rs <- dbExecute(dbcon, sql)
```
```{sql connection=dbcon}
SELECT * FROM Faculty
```
Populating Department Table
```{r}
get.dep <- "select distinct DepartmentCode,DepartmentName, DepartmentHead from `stu.data.raw`"
dep.info <- sqldf(get.dep)
nrows <- nrow(dep.info)
sql <- "insert into Departments (department_id, department_name, department_head, department_code) values"

for (i in 1:nrows) {
  id <- i
  name <- dep.info$DepartmentName[i]
  code <- dep.info$DepartmentCode[i]
  head <- dep.info$DepartmentHead[i]
  if (head == -1) {
    head <- "NULL"    
  }
  val = paste0("(", id, ",'", name, "',", head, ",'", code, "')")
  if (i != nrows) {
    val = paste0(val, ",")
  }
  sql <- paste0(sql, val)
}
rs <- dbExecute(dbcon, sql)
```
```{sql connection=dbcon}
SELECT * FROM Departments
```
Populate Department Members
```{r}
# probably need to go through the list and select distinct faculty again then get the dep
sql <- "select distinct (FacultyID), DepartmentCode, DepartmentName from `stu.data.raw` order by facultyid"
fd.info <- sqldf(sql) # There is an issue with faculty members teaching in various departments - resolve this
sql <- "insert into DepartmentMembership (dm_id, faculty_id, department_id) values "
nrows <- nrow(fd.info)
for (i in 1:nrows) {
  dept.id <- dbGetQuery(dbcon, paste0("select department_id from departments where department_code = '", fd.info$DepartmentCode[i], "'" ))
  val <- paste0("(", i, ",", fd.info$FacultyID[i], ",", dept.id, ")")
  if (i != nrows) {
    val = paste0(val, ",")
  }
  sql <- paste0(sql, val)
}

rs <- dbExecute(dbcon, sql)
```
```{sql connection=dbcon}
SELECT * FROM DepartmentMembership
```

Populate Courses Table:
```{r}
get.courses <- "select distinct coursecode, semester, departmentcode, facultyid, CourseName from `stu.data.raw`"
course.info <- sqldf(get.courses)
nrows <- nrow(course.info)
sql <- "insert into Courses (course_id, course_name, department_id, faculty_id, semester_offered, course_code) values"

for (i in 1:nrows) {
  id <- i
  course.code <- course.info$CourseCode[i]
  course.name <- course.info$CourseName[i]
  semester <- course.info$Semester[i]
  department.code <- course.info$DepartmentCode[i]
  fid <- course.info$FacultyID[i]
  id.sql <- paste0("select department_id from departments where department_code = '", department.code, "'")
  dep.id <- dbGetQuery(dbcon, id.sql)
  val = paste0("(", id, ",'", course.name, "',", dep.id, ",", fid, ",'", semester, "','", course.code, "')")
  if (i != nrows) {
    val = paste0(val, ",")
  }
  sql <- paste0(sql, val)
}
rs <- dbExecute(dbcon, sql)
```
```{sql connection=dbcon}
SELECT * FROM Courses
```
Populating Classrooms Table
```{r}
get.rooms <- "select distinct Building, RoomNumber, Capacity from `stu.data.raw`"
rooms.info <- sqldf(get.rooms)
rooms.info
nrows <- nrow(rooms.info)
sql <- "insert into Classrooms (classroom_id, room_number, building_name, capacity) values"

for (i in 1:nrows) {
  id <- i
  room.num <- rooms.info$RoomNumber[i]
  building <- rooms.info$Building[i]
  capacity <- rooms.info$Capacity[i]
  val = paste0("(", id, ",", room.num, ",'", building, "',", capacity, ")")
  if (i != nrows) {
    val = paste0(val, ",")
  }
  sql <- paste0(sql, val)
}

rs <- dbExecute(dbcon, sql)
```
```{sql connection=dbcon}
SELECT * FROM Classrooms ORDER BY building_name DESC
```
Populating Course Schedules and Course Schedule Days Junction Table:
```{r}
get.courses <- "select distinct coursecode, coursename, semester, departmentcode, Building, RoomNumber, Capacity, DaysOfWeek, StartTime,EndTime from `stu.data.raw`"
c.info <- sqldf(get.courses)
nrows = nrow(c.info)

sql <- "insert into CourseSchedules (schedule_id, course_id, classroom_id, start_time, end_time) values "
csd.sql <- "insert into CourseScheduleDays (schedule_id, day_id) values "
for (i in 1:nrows) {
  b = c.info$Building[i]
  r = c.info$RoomNumber[i]
  capacity = c.info$Capacity[i]
  get.cid <- "select classroom_id from classrooms where building_name = '"
  get.cid <- paste0(get.cid, b, "' and room_number = ", r, " and capacity = ", capacity)
  room.id = dbGetQuery(dbcon, get.cid)
  course.id <- dbGetQuery(dbcon, paste0("select c.course_id, d.department_code from courses c, departments d  where d.department_id = c.department_id and c.course_code = '", c.info$CourseCode[i], "' and d.department_code = '", c.info$DepartmentCode[i], "' and c.semester_offered = '", c.info$Semester[i], "'"))
  val = paste0("(", i, ",", course.id$course_id, ",", room.id, ",'", c.info$StartTime[i], "','", c.info$EndTime[i],"'", ")")
  if (i != nrows) {
    val = paste0(val, ",")
  }
  sql <- paste0(sql, val)
  days.split <- strsplit(c.info$DaysOfWeek[i], split="", fixed=T)
  for (j in 1:length(days.split)) {
    get.day.id <- dbGetQuery(dbcon, paste0("select day_id from DaysOfWeek where day_name = '", days.split[[1]][j], "'"))
    days.val <- paste0("(", i, ",", get.day.id$day_id, ")")
    if (i == nrows) {
      if (j == length(days.split)) {
        csd.sql <- paste0(csd.sql, days.val)
      }
    } else {
      csd.sql <- paste0(csd.sql, days.val, ",")
    }
  }
}

rs <- dbExecute(dbcon, sql)
rs <- dbExecute(dbcon, csd.sql)
```

```{sql connection=dbcon}
select * from courseScheduleDays
```

Populating Major Table and StudentMajor Table
```{r}
major.info <- "select distinct StudentID, StudentMajor from `stu.data.raw`"
major.info <- sqldf(major.info)
majors <- unique(major.info$StudentMajor)
m.rows <- length(majors[])
major.sql <- "insert into majors (major_name, department_id) values "
for (i in 1:m.rows) {
  dep.id <- dbGetQuery(dbcon, paste0("select department_id from departments where department_name ='", majors[i], "'"))
  val <- paste0("('", majors[i], "',", dep.id, ")")
  if (i != m.rows) {
    val <- paste0(val, ",")
  }
  major.sql <- paste0(major.sql, val)
}

rs <- dbExecute(dbcon, major.sql)

sql <- "insert into StudentMajors (student_id, major_id) values "
for (i in 1:nrow(major.info)) {
  major.id <- dbGetQuery(dbcon, paste0("select major_id from majors where major_name = '", major.info$StudentMajor[i], "'"))
  val <- paste0 ("(", major.info$StudentID[i], ",", major.id, ")")
  if (i != nrow(major.info)) {
    val <- paste0(val, ",")
  }
  sql <- paste0(sql, val)
}

rs <- dbExecute(dbcon, sql)
```
```{sql connection=dbcon}
SELECT * FROM StudentMajors
```

```{r}
enrollment.info <- "select distinct StudentID, CourseCode, Semester, DepartmentCode, Grade from `stu.data.raw`"
enrollment.info <- sqldf(enrollment.info)
enrollment.info
sql <- "insert into enrollment (student_id, course_id, grade) values "
for (i in 1:nrow(enrollment.info)) {
  stu.id <- enrollment.info$StudentID[i]
  #print(stu.id)
  c.code <- enrollment.info$CourseCode[i]
  dep.id <- dbGetQuery(dbcon, paste0("select department_id from departments where department_code = '", enrollment.info$DepartmentCode[i], "'"))
  c.id <- dbGetQuery(dbcon, paste0("select course_id from courses where course_code = '", c.code, "' and semester_offered ='", enrollment.info$Semester[i], "' and department_id ='", dep.id, "'"))
  grade <- enrollment.info$Grade[i]
  if (is.na(grade)) {
    grade <- "NULL"
    val <- paste0("(", stu.id, ",", c.id, ",", grade, ")")
  }
  else {
    val <- paste0("(", stu.id, ",", c.id, ",", grade, ")")
  }
  if (i != nrow(enrollment.info)) {
    val <- paste0(val, ",")
  }
  sql <- paste0(sql, val)
}
#print(sql)
rs <- dbExecute(dbcon, sql)
```
```{sql connection=dbcon}
SELECT * FROM ENROLLMENT
```


```{sql connection=dbcon}
SELECT * FROM CoursesWithComposite
```

## Making Some Sample Queries
First, get the average grade for all students at the university
```{sql connection=dbcon}
SELECT avg(e.grade) from enrollment e
```
Get all of the grades for each student
```{sql connection=dbcon}
select distinct s.student_id, e.grade from students s, enrollment e where s.student_id = e.student_id
```

Get the average grade for each student
```{sql connection=dbcon}
select student_id, avg(grade) as AvgGrade from enrollment group by student_id
```
Get students with an average grade above the average of the university
```{sql connection=dbcon}
WITH avgGrades as (
select student_id, avg(grade) as AvgGrade from enrollment group by student_id
),
uniAvg as (
SELECT avg(e.grade) as UniAvg from enrollment e
)
select a.student_id, a.AvgGrade from avgGrades a where (a.AvgGrade > (select UniAvg from uniAvg)) ORDER BY a.AvgGrade DESC
```

Get students who qualify for cum laude:

Get students who qualify for magna cum laude:

Get students who qualift for summa cum laude:
## Adding Some Triggers 

## Inserting New Data and Checking Triggers

## Finish Up

Finally, disconnect from the database
```{r}
dbDisconnect(dbcon)
```