---
title: "Example University DB Implementation"
author: "Raphael Russo"
output: html_notebook
---

## Overview
This is a simple example implementation of a relational database for a university based off of the schema outlined in `University_ERD.pdf` which is in BCNF. It includes students, majors, courses, faculty, etc. Data will be loaded a csv file of students and the courses they have taken, and stored in a SQLite database using R and SQL, and at the end, various queries will be made. Student data is generated using `generate_data.py` and can be found in `generated_u_data.csv`.

## Connect To Database

First, connect to the database, or create a new one if it does not already exist. 
```{r}
library(RSQLite)
library(DBI)
library(sqldf)
library(reticulate)
data.file <- "generated_u_data.csv"
dbfile <- "UniDB.db"
if (file.exists(dbfile)) {
  file.remove(dbfile) # just get rid of the db instead of dropping all the tables to start fresh
}
if (file.exists(data.file)) {
  file.remove(data.file)
}
py_run_file("generate_data.py")
dbcon <- dbConnect(RSQLite::SQLite(), dbfile)
dbExecute(dbcon, "PRAGMA foreign_keys = ON")
```

## Create Tables

Students Table:
```{sql createStudents, connection=dbcon}
CREATE TABLE IF NOT EXISTS Students (
  student_id INTEGER,
  first_name TEXT NOT NULL,
  last_name TEXT,
  birth_date DATE NOT NULL,
  email TEXT,
  expected_graduation DATE,
  PRIMARY KEY (student_id)
)
```

Faculty Table:
```{sql createFaculty, connection=dbcon}
CREATE TABLE IF NOT EXISTS Faculty (
  faculty_id INTEGER,
  first_name TEXT NOT NULL,
  last_name TEXT,
  birth_date DATE NOT NULL,
  email TEXT,
  PRIMARY KEY (faculty_id)
)
```

Departments Table:
TO DO - ADD DEPARTMENT HEAD IN DATA GEN
```{sql createDepartments, connection=dbcon}
CREATE TABLE IF NOT EXISTS Departments (
  department_id INTEGER,
  department_name TEXT NOT NULL,
  department_head INTEGER,
  department_code TEXT NOT NULL,
  PRIMARY KEY (department_id),
  FOREIGN KEY (department_head) REFERENCES Faculty (faculty_id) 
)
```

Department Members Table:
```{sql createDepartmentMembers, connection=dbcon}
CREATE TABLE IF NOT EXISTS DepartmentMembers (
  dm_id INTEGER,
  faculty_id INTEGER NOT NULL,
  department_id INTEGER NOT NULL,
  PRIMARY KEY (dm_id),
  FOREIGN KEY (faculty_id) REFERENCES Faculty (faculty_id),
  FOREIGN KEY (department_id) REFERENCES Departments (department_id)
)
```

Courses Table:
```{sql createCourses, connection=dbcon}
CREATE TABLE IF NOT EXISTS Courses (
  course_id INTEGER,
  course_name TEXT NOT NULL,
  course_code INTEGER NOT NULL,
  department_id INTEGER NOT NULL,
  faculty_id INTEGER NOT NULL,
  semester_offered TEXT NOT NULL,
  PRIMARY KEY (course_id),
  FOREIGN KEY (department_id) REFERENCES Departments,
  FOREIGN KEY (faculty_id) REFERENCES Faculty (faculty_id)
)
```

Composite Key for Courses:
```{sql createCompositeCourseKeyView, connection=dbcon}
CREATE VIEW CoursesWithComposite AS
SELECT d.department_code || c.course_code || c.semester_offered AS composite_key
FROM Courses c, Departments d where c.department_id = d.department_id
```

Majors Table:
```{sql createMajors, connection=dbcon}
CREATE TABLE IF NOT EXISTS Majors (
  major_id INTEGER,
  major_name TEXT NOT NULL,
  department_id INTEGER NOT NULL,
  PRIMARY KEY (major_id),
  FOREIGN KEY (department_id) REFERENCES Departments (department_id)
)
```

Student Majors Table:
```{sql creatStudentMajors, connection=dbcon}
CREATE TABLE IF NOT EXISTS StudentMajors (
  student_id INTEGER NOT NULL,
  major_id INTEGER NOT NULL,
  CONSTRAINT student_major_id PRIMARY KEY (student_id, major_id),
  FOREIGN KEY (student_id) REFERENCES Students (student_id),
  FOREIGN KEY (major_id) REFERENCES Majors (major_id)
  
)
```


Required Courses Table:
```{sql createRequiredCourses, connection=dbcon}
CREATE TABLE IF NOT EXISTS RequiredCourses (
  rc_id INTEGER,
  course_id INTEGER NOT NULL,
  major_id INTEGER NOT NULL,
  PRIMARY KEY (rc_id),
  FOREIGN KEY (course_id) REFERENCES Courses (course_id),
  FOREIGN KEY (major_id) REFERENCES Majors (major_id)
)
```

Elective Courses Table:
```{sql createElectiveCourses, connection=dbcon}
CREATE TABLE IF NOT EXISTS ElectiveCourses (
  ec_id INTEGER,
  course_id INTEGER NOT NULL,
  major_id INTEGER NOT NULL,
  PRIMARY KEY (ec_id),
  FOREIGN KEY (course_id) REFERENCES Courses (course_id),
  FOREIGN KEY (major_id) REFERENCES Majors (major_id)
)
```

Classrooms Table:
```{sql createClassrooms, connection=dbcon}
CREATE TABLE IF NOT EXISTS Classrooms (
  classroom_id INTEGER,
  room_number INTEGER,
  building_name TEXT,
  capacity INTEGER,
  PRIMARY KEY (classroom_id)
)
```

Course Schedules Table:
```{sql createCourseSchedules, connection=dbcon}
CREATE TABLE IF NOT EXISTS CourseSchedules (
  schedule_id INTEGER,
  course_id INTEGER NOT NULL,
  classroom_id INTEGER NOT NULL,
  start_time TIME,
  end_time TIME,
  PRIMARY KEY (schedule_id)
)
```

Days of Week Table:
```{sql createDaysOfWeek, connection=dbcon}
CREATE TABLE IF NOT EXISTS DaysOfWeek (
    day_id INTEGER PRIMARY KEY,
    day_name TEXT(1) NOT NULL
);

```

Course Scheduled Days Table:
```{sql createScheduledDays, connection=dbcon}
CREATE TABLE IF NOT EXISTS CourseScheduleDays (
    schedule_id INTEGER,
    day_id INTEGER,
    CONSTRAINT cs_id PRIMARY KEY (schedule_id, day_id),
    FOREIGN KEY (schedule_id) REFERENCES CourseSchedules(schedule_id),
    FOREIGN KEY (day_id) REFERENCES DaysOfWeek(day_id)
);
```

## Check Table Creation

## Loading CSV Data

Read the student data csv:
```{r loadStudentData}
stu.data.raw <- read.csv("generated_u_data.csv")
```

Populate Days of Week table:
```{r insertDaysOfWeek}
days = c("U", "M", "T", "W", "R", "F", "S")
ndays = length(days)
sql <- "insert into DaysOfWeek (day_id, day_name) values"
for (i in 1:ndays) {
  day = days[i]
  val = paste0("(", i, ",'", day, "')")
  if (i != ndays) {
    val = paste0(val, ",")
  }
  sql <- paste0(sql, val)
}

print(sql)
rs <- dbExecute(dbcon, sql)
```
```{sql connection=dbcon}
SELECT * FROM DaysOfWeek
```
Populate Student Table
```{r}
get.students <- "select distinct StudentID,StudentFirstName,StudentLastName,StudentEmail,StudentDOB,StudentGradDate from `stu.data.raw`"
stu.info <- sqldf(get.students)
nrows <- nrow(stu.info)
sql <- "insert into Students (student_id, first_name, last_name, birth_date, email, expected_graduation) values"

for (i in 1:nrows) {
  id <- stu.info$StudentID[i]
  f.name <- stu.info$StudentFirstName[i]
  l.name <- stu.info$StudentLastName[i]
  email <- stu.info$StudentEmail[i]
  dob <- stu.info$StudentDOB[i]
  grad <- stu.info$StudentGradDate[i]
  val = paste0("(", id, ",'", f.name, "','", l.name, "','", dob, "','", email, "','", grad, "')")
  if (i != nrows) {
    val = paste0(val, ",")
  }
  sql <- paste0(sql, val)
}

rs <- dbExecute(dbcon, sql)
```
```{sql connection=dbcon}
SELECT * FROM STUDENTS
```
Populate Faculty Table
```{r}
get.fac <- "select distinct FacultyID,FacultyFirstName,FacultyLastName,FacultyDOB, FacultyEmail from `stu.data.raw`"
fac.info <- sqldf(get.fac)
nrows <- nrow(fac.info)
sql <- "insert into Faculty (faculty_id, first_name, last_name, birth_date, email) values"


for (i in 1:nrows) {
  id <- fac.info$FacultyID[i]
  f.name <- fac.info$FacultyFirstName[i]
  l.name <- fac.info$FacultyLastName[i]
  email <- fac.info$FacultyEmail[i]
  dob <- fac.info$FacultyDOB[i]
  val = paste0("(", id, ",'", f.name, "','", l.name, "','", dob, "','", email, "')")
  if (i != nrows) {
    val = paste0(val, ",")
  }
  sql <- paste0(sql, val)
}
rs <- dbExecute(dbcon, sql)
```
```{sql connection=dbcon}
SELECT * FROM Faculty
```
Populating Department Table
```{r}
get.dep <- "select distinct DepartmentCode,DepartmentName, DepartmentHead from `stu.data.raw`"
dep.info <- sqldf(get.dep)
nrows <- nrow(dep.info)
sql <- "insert into Departments (department_id, department_name, department_head, department_code) values"

for (i in 1:nrows) {
  id <- i
  name <- dep.info$DepartmentName[i]
  code <- dep.info$DepartmentCode[i]
  head <- dep.info$DepartmentHead[i]
  if (head == -1) {
    head <- "NULL"    
  }
  val = paste0("(", id, ",'", name, "',", head, ",'", code, "')")
  if (i != nrows) {
    val = paste0(val, ",")
  }
  sql <- paste0(sql, val)
}
rs <- dbExecute(dbcon, sql)
```
```{sql connection=dbcon}
SELECT * FROM Departments
```
Populate Department Members
```{r}
# probably need to go through the list and select distinct faculty again then get the dep
sql <- "select distinct (FacultyID), DepartmentCode, DepartmentName from `stu.data.raw` order by facultyid"
sqldf(sql) # There is an issue with faculty members teaching in various departments - resolve this
```
Populate Courses Table:
```{r}
get.courses <- "select distinct coursecode, semester, departmentcode, facultyid, CourseName from `stu.data.raw`"
course.info <- sqldf(get.courses)
nrows <- nrow(course.info)
sql <- "insert into Courses (course_id, course_name, department_id, faculty_id, semester_offered, course_code) values"

for (i in 1:nrows) {
  id <- i
  course.code <- course.info$CourseCode[i]
  course.name <- course.info$CourseName[i]
  semester <- course.info$Semester[i]
  department.code <- course.info$DepartmentCode[i]
  fid <- course.info$FacultyID[i]
  id.sql <- paste0("select department_id from departments where department_code = '", department.code, "'")
  dep.id <- dbGetQuery(dbcon, id.sql)
  val = paste0("(", id, ",'", course.name, "',", dep.id, ",", fid, ",'", semester, "','", course.code, "')")
  if (i != nrows) {
    val = paste0(val, ",")
  }
  sql <- paste0(sql, val)
}
rs <- dbExecute(dbcon, sql)
```
```{sql connection=dbcon}
SELECT * FROM Courses
```
Populating Classroom Table
```{r}
get.rooms <- "select distinct Building, RoomNumber, Capacity from `stu.data.raw`"
rooms.info <- sqldf(get.rooms)
rooms.info
nrows <- nrow(rooms.info)
sql <- "insert into Classrooms (classroom_id, room_number, building_name, capacity) values"

for (i in 1:nrows) {
  id <- i
  room.num <- rooms.info$RoomNumber[i]
  building <- rooms.info$Building[i]
  capacity <- rooms.info$Capacity[i]
  val = paste0("(", id, ",", room.num, ",'", building, "',", capacity, ")")
  if (i != nrows) {
    val = paste0(val, ",")
  }
  sql <- paste0(sql, val)
}

rs <- dbExecute(dbcon, sql)
```
```{sql connection=dbcon}
SELECT * FROM Classrooms ORDER BY building_name DESC
```
Populating Course Schedules Table:
```{r}
get.courses <- "select distinct coursecode, coursename, semester, departmentcode, Building, RoomNumber, Capacity, DaysOfWeek, StartTime,EndTime from `stu.data.raw`"
c.info <- sqldf(get.courses)
nrows = nrow(c.info)

sql <- "insert into CourseSchedules (schedule_id, course_id, classroom_id, start_time, end_time) values "
for (i in 1:nrows) {
  b = c.info$Building[i]
  r = c.info$RoomNumber[i]
  capacity = c.info$Capacity[i]
  get.cid <- "select classroom_id from classrooms where building_name = '"
  get.cid <- paste0(get.cid, b, "' and room_number = ", r, " and capacity = ", capacity)
  room.id = dbGetQuery(dbcon, get.cid)
  course.id <- dbGetQuery(dbcon, paste0("select c.course_id, d.department_code from courses c, departments d  where d.department_id = c.department_id and c.course_code = '", c.info$CourseCode[i], "' and d.department_code = '", c.info$DepartmentCode[i], "' and c.semester_offered = '", c.info$Semester[i], "'"))
  val = paste0("(", i, ",", course.id$course_id, ",", room.id, ",'", c.info$StartTime[i], "','", c.info$EndTime[i],"'", ")")
  if (i != nrows) {
    val = paste0(val, ",")
  }
  sql <- paste0(sql, val)
}

rs <- dbExecute(dbcon, sql)
```

```{sql connection=dbcon}
select * from courseSchedules
```


```{sql connection=dbcon}
SELECT * FROM CoursesWithComposite
```

## Making Some Sample Queries

## Adding Some Triggers 

## Inserting New Data and Checking Triggers

## Finish Up

Finally, disconnect from the database
```{r}
dbDisconnect(dbcon)
```